Bluetoothを用いたマルチVM対応\\mrubyバイトコードローダ

### 2016-02-17 ETNET2016/thesis.tex
ETNET2016 first

%        File: thesis.tex
%     Created: Tue Feb 16 06:00 PM 2016 J
% Last Change: Tue Feb 16 06:00 PM 2016 J
%
\input{../dummy/dummy.tex}
\documentclass[submit]{ipsj_v2/UTF8/ipsj}
%\documentclass{ipsj}

@@

\etitle{mruby Bytecode Loader Using Bluetooth in Multi-VM Environment}

\affiliate{IPSJ}{情報処理学会\\
IPSJ, Chiyoda, Tokyo 101--0062, Japan}

\paffiliate{JU}{情報処理大学\\
Johoshori University}

\author{山本 拓朗}{Takuro Yamamoto}{IPSJ}[joho.taro@ipsj.or.jp]
\author{大山 博司}{Hiroshi Oyama}{IPSJ}
\author{安積 卓也}{Takuya Azumi}{IPSJ,JU}[gakkai.jiro@ipsj.or.jp]

@@

\begin{jkeyword}
情報処理学会論文誌ジャーナル，\LaTeX，スタイルファイル，べからず集
\end{jkeyword}

\begin{eabstract}
This document is a guide to prepare a draft for submitting to IPSJ
Journal, and the final camera-ready manuscript of a paper to appear in
IPSJ Journal, using {\LaTeX} and special style files.  Since this
document itself is produced with the style files, it will help you to
refer its source file which is distributed with the style files.
\end{eabstract}

\begin{ekeyword}
IPSJ Journal, \LaTeX, style files, ``Dos and Don'ts'' list
\end{ekeyword}

@@

% Contribution: The proposed framework gives the contribution in the following points.
% 1) To improve the software development efficiency:
% Developers do not need to rewrite a storage/ROM device and also to restart an OS. 
% The loader supports the countinuous loading, which saves the Bluetooth setup time. 
% Therefore, the mruby bytecode loader using Bluetooth helps developers develop software.
% 2) To effectively execute multiple mruby programs in concurrent or/and parallel: 
% Developers can execute multiple tasks without the knowledge of an RTOS because the RiteVM sceduler switches tasks cyclically.
% In additon, synchronization of multiple tasks is also implemented.
% 3) To focus on the benefits of component-based developments:
% This paper shows the specific examples for the benefits of component-based development.

@@

\section{既存フレームワーク}
\label{fig:background}
図\ref{fig:proposed}に提案フレームワークのシステムモデルを示す．
RiteVMとmrubyライブラリを含むプラットフォーム部分は，はじめにターゲットデバイス上で起動されている．
mrubyアプリケーションのバイトコードは，ホストからターゲットデバイスへと転送される．
それぞれバイトコードは，RiteVMに割り当てられ，並行して起動される．

@@

\subsubsection{Component Model}
図\ref{fig:component}にコンポーネント図の例を示す．
{\it cell} which is an instance of a component in TECS consists of {\it entry} ports, {\it call} ports, attributes and internal variables.

A {\it entry} port is an interface to provide functions with other {\it cell}s, and a {\it call} port is an interface to use functions of other {\it cell}s.
A {\it cell} has one or more {\it entry} ports and {\it call} ports.
Functions of a {\it cell} are implemented in the C language.

A type of a {\it entry}/{\it call} port is defined by a {\it signature} which is a set of functions.
A {\it signature} is the interface definition of a {\it cell}.
The {\it call} port of a {\it cell} can be connected to the {\it entry} port of another {\it cell} with the same {\it signature}.
A {\it celltype} is the definition of a {\it cell}.
{\it Celltype} defines one or more {\it call}/{\it entry} ports, attributes and internal variables.

@@

\subsubsection{Component Description}
The description of a component in TECS is divided into three parts, {\it signature}, {\it celltype}, and build description.
TECS code is written in .cdl (component description language) file.
The component description is mentioned with an example shown in Figure \ref{fig:component} as follow.

\begin{description}
    \item[{\bf Signature Description}]\mbox{}\\
        The {\it signature} description defines an interface of a {\it cell}.
        A {\it signature} name is described following the keyword {\it signature}.
        It also has the prefix ``s".
        In this way, a {\it signature} is defined such as sMotor shown in Figure \ref{signature}.
        To make the definition of an interface clear, specifiers such as in and out are used in TECS.
        [in] and [out] represent input and output, respectively.\\
\begin{figure}[t]
\centering
\begin{lstlisting}
signature sMotor{
    int32_t getCounts( void );
    ER resetCounts( void );
    ER setPower( [in]int power );
    ER stop( [in] bool_t brake );
    ER rotate( [in] int degrees, [in] uint32_t speed_abs,
              [in] bool_t blocking );
    void initializePort( [in]int32_t type );
};
\end{lstlisting}
\caption{Signature Description}
\label{signature}
\end{figure}

    \item[{\bf Celltype Description}]\mbox{}\\
        The {\it celltype} description defines {\it entry} ports, {\it call} ports, attributes, and valuables of a {\it celltype}.
        An example of a {\it celltype} description is shown in Figure \ref{celltype}.
        A {\it celltype} name following the keyword {\it celltype} with the prefix ``t" and elements of a {\it celltype} is described.
        To define {\it entry} ports, a {\it signature} such as sMotor, and an {\it entry} port name such as eMotor follow the keyword {\it entry}.
        In the same way, {\it call} ports can be declared.
        Attributes and valuables follow the keyword {\it attr} and {\it var} respectively.\\
\begin{figure}[t]
\centering
\begin{lstlisting}
celltype tCaller{
    call sMotor cMotor;
};
celltype tMotor{
    entry sMotor eMotor;
    attr{
        int32_t attr = 100;
    };
    var{
        int32_t var;
    };
};
\end{lstlisting}
\caption{Celltype Description}
\label{celltype}
\end{figure}

    \item[{\bf Build Description}]\mbox{}\\
        The build description is used to instantiate {\it cell}s and connect {\it cell}s.
        Figure \ref{build} shows an example of a build description.
        A {\it celltype} name such as tMotor, and a {\it cell} name such as Motor follow the keyword {\it cell}.
        To compose {\it cell}s, a {\it call} port, a {\it signature}, a {\it entry} port in order are described.
        In this example, a {\it entry} port eMotor in a {\it cell} Motor is connected to a {\it call} port cMotor in a {\it cell} Caller.\\
\begin{figure}[t]
\centering
\begin{lstlisting}
cell tMotor Motor{
};
cell tCaller Caller{
    cMotor = Motor.eMotor;
};
\end{lstlisting}
\caption{Build Description}
\label{build}
\end{figure}

@@

\subsection{mruby on TECS}
\label{sec:mruby on TECS}
mruby on TECS is a component-based framework for running script language.
This framework uses two technologies, mruby and TECS.

\subsubsection{System Model}\mbox{}\\

The system model of mruby on TECS is shown in Figure \ref{fig:mrubyontecs}.
Each mruby program, which is bytecode, runs on its own RiteVM as a componentized task of an RTOS.
TECS components support various embedded drivers such as motor and sensor drivers.

An mruby-TECS bridge plays a role to call a native program (e.g. C legacy code) from an mruby program.
The mruby-TECS bridge provides native libraries for mruby.
It also gives TECS components to receive the invocation from an mruby program.
The mruby-TECS bridge is described in more detail bellow.

As the target RTOS, TOPPERS/HRP2 \cite{url:HRP2}, \cite{par:hr-tecs} is used in this paper.
TOPPERS/HRP2 is an RTOS based on $\mu$ITRON \cite{par:microITRON} with memory protection.
However, mruby on TECS does not depend on the RTOS because TECS supports not only TOPPERS/HRP2 but also the other RTOSs such as OSEK \cite{par:OSEK} and TOPPERS/ASP \cite{url:ASP}.

@@

\subsubsection{mruby-TECS Bridge}\mbox{}\\

There is a great difference between the execution time of mruby and C language.
According to  \cite{par:mrubyonTECS}, mruby programs are several hundreds times slower than C programs.
The execution of mruby bytecode on a RiteVM is not as efficient as that of C language.
Thus it is difficult to use mruby for all of code.

The use of Ruby on embedded devices provides the benefit of productivity and maintainability due to the ease to use and read.
On the other hands, it is necessary to implement parts of applications in C language in order to manipulate actuators and sensors, and also make a critical section of code run quickly.

Figure \ref{fig:mruby_TECS_bridge} shows an example of use of an mruby-TECS bridge for controlling a motor.
The left side of BridgeMotor belongs to the mruby program.
The right side of BridgeMotor belongs to TECS component.
\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/mruby_TECS_bridge.pdf}
    \caption{mruby-TECS Bridge}
    \label{fig:mruby_TECS_bridge}
\end{figure}

The mruby-TECS bridge generates two things.
One is a {\it celltype} to receive invocation from the mruby program.
The other thing is an mruby class that corresponds to the TECS component specified by the developers to invoke a C function from the mruby program.

A code of an mruby-TECS bridge is generated.
The generation code supports registration of classes and methods for mruby.
The methods in an mruby class are defined by generation codes for an mruby-TECS bridge, such as setPower and stop.
Thus, when a method is called in an mruby program, an mruby-TECS bridge calls the function defined in the TECS component such as a Motor {\it cell}.

\section{Design and Implementation}
\label{sec:Design and Implementation}
図\ref{fig:system_model}に提案フレームワークの詳細なシステムモデルを示す．
ホストから転送されたmrubyアプリケーションのバイトコードは，それぞれのRiteVMに実装されたローダで受信され，同期処理により同時に実行される．
RiteVMスケジューラがタスクを周期的に切り替えるため，mrubyアプリケーションは並行動作することができる．
　
Figure \ref{fig:system_model} shows the detailed system model of the proposed framework.
Each mruby application bytecode transferred from the host is received by the loader in the RiteVM.
The RiteVM reads the own bytecode and executes it.
The mruby applications run at the same time because of synchronization processing.
The RiteVM scheuler switches tasks as multiple tasks can run in concurrent.

\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/system_model.pdf}
    \caption{Detailed System Model of the proposed framework}
    \label{fig:system_model}
\end{figure}

\subsection{mruby Bytecode Loader Using Bluetooth}
\label{sec:mruby bytecode loader using Bluetooth}
この章では，Bluetoothを用いたmrubyバイトコードローダについて述べる．
現状のmruby on TECSでは，mrubyバイトコードをプラットフォームに組み込んでいるため，mrubyプログラムを修正する度にプラットフォーム部分をもう一度コンパイル・リンクし，SDカードやROMといったデバイスに再度書き込み，OSを再起動する必要がある．
このような作業を繰り返すことで，作業効率は悪くなる．
提案フレームワークでは，プラットフォーム部分をコンパイル・リンクし記憶デバイスに書き込むのは，はじめの一度だけで良いため，作業効率を向上させることができる．

@@

\subsubsection{Component of RiteVM with mruby bytecode loader using Bluetooth}
提案フレームワークでは，mrubyバイトコードローダはTECSのコンポーネントとして提供されている．
このコンポーネントは，RiteVMのコンポーネントの拡張として実装されている．(詳細については\cite{par:mrubyonTECS}を参照)
このコンポーネントでは，転送されたバイトコードの受信に加え，RiteVMのコンフィグレーションが行われる．

@@

\end{lstlisting}
\caption{Main code for RiteVM with mruby bytecode loader using Bluetooth}
\label{maincode_mrubybluetooth}
\end{figure}
\subsection{Multitask}
\label{sec:Multitask}
この章では，提案フレームワークでのマルチタスク処理の設計にちて述べる．
mruby on TECSは，マルチタスクに対応しているが，開発者がRTOSの知識を熟知している必要がある．

@@

\subsubsection{RiteVM Scheduler}
RiteVMスケジューラは，周期ハンドラであり，μITRONのサービスコールである{\it rotateReadyQueue}を呼び出す．
{\it rotateReadyQueue}は，同優先度のタスクの実行を切り替える関数である．
RiteVMスケジューラの設計を図\ref{fig:rotateReadyQueue}に示す．

@@

\subsubsection{Component of RiteVM Scheduler}
図\ref{fig:cyclic_handler}にRiteVMスケジューラのコンポーネント図を示す．
RiteVMスケジューラのコンポーネントは，CyclicHandlerとCyclicMainから構成される．
CyclicHandlerセルは，μITRONの周期ハンドラの設定を行う．
μITRONの周期ハンドラは，\cite{par:microITRON}に詳しく書かれている．

@@

図\ref{build_cyclic_handler}に，図\ref{fig:cyclic_handler}に示されるコンポーネントの組み上げ記述を示す．

@@

\subsubsection{Synchronization of Multiple RiteVM Tasks}
複数のmrubyアプリケーションの起動を同期させるために，提案フレームワークでは，同期処理の手法としてイベントフラグを適用した．
タスクはそれぞれ0x01(01)，0x02(10)のパターンをセットし，待ちパターン0x3(11)でAND待ちする．
この機構は，2つ以上のタスクの場合でも適用できる．
例えば，タスクが4つの場合，タスクはそれぞれ，0x01(0001), 0x02(0010), 0x04 (0100), 0x08 (1000)のパターンをセットし，待ちパターン0x0f(1111) でAND待ちする．(図\ref{fig:Eventflag} (A))

@@

\subsection{Benefits of Component-Based Development}
この章では，コンポーネントベース開発の利点を使った設計について述べる．

@@

\section{Experimental Evaluation}
\label{sec:Evaluation}
This section mentions experimental results and their consideration.
To analyze the advantages of the proposed framework, the evaluations are performed as follows.
\begin{itemize}
%    \item Execution time of the platform
    \item Size and time for transferred mruby bytecodes
    \item Execution time with singletasking, co-routine, and multitasking
    \item Overhead for cyclic time
%    \item Synchronization of mruby applications
    \item Code size using the benefit of CBD 
\end{itemize}

These evaluations are performed in order to indicate that an mruby bytecode loader improves the software development efficiency, and that the proposed multitask processing effectively executes compared with singletasking or co-routine, and also the overhead of the cyclic period.
This paper demonstrates the proposed system on a LEGO MINDSTORMS EV3 (300MHz ARM9-based Sitara AM1808 system-on-a-chip) compiled with gcc 4.9.3 -O2 and mruby version 1.2.0.

The size, load process time, and compilation time for mruby application and mruby application including library is shown in Table \ref{tab:size_and_time}.
The overhead of load processing is 50.933 msec, which it takes to load a zero bytes bytecode.
Similarly, the overhead of compilation is 46.9 msec, which it takes to compile a zero bytes program.
The mruby application bytecode is smaler and faster than that of including mruby libraries in all terms.
The difference becomes larger as the number of RiteVMs increases because it takes the 50 msec overhead per one RiteVM. 
%In the proposed framework, developers send only mruby application and prepare mruby library in advance.
%Therefore, the design can send the bytecode faster.
In addition, the design can save the time to connect a storage/ROM device to restart an OS.
These advantages lead improvement of the software development efficiency.

@@

The comparison of the application execution time with singletasking, co-routine, and multitasking is shown in Figure \ref{fig:comparison_s_c_m}.
The 100,000 times loop program is used as mruby application for evaluation of execution time.
In detail, the singletask program loops 100,000 times, and the multitask and co-routine programs loop 50,000 times in each task.
In Figure \ref{fig:comparison_s_c_m}, the cyclic time of the cyclic handler for multitasking is one msec.
While multitask's execution time is slower than singletask's, co-routine takes more time than multitask to execute an mruby application.
This result shows the proposed design is superior to co-routine in terms of execution time.
The overhead of multitask is about 5 \% ($(multitask-singletask)/multitask$).
Switching tasks' overhead is also evaluated, that is about three $\mu$sec on average.
%The number of switching tasks is from 500 to 600 because singletasking process in Figure \ref{fig:comparison_s_c_m} takes about 540 msec.
%Therefore, the overhead of multitasking is 0.5 \% or less, and multitasking can be used without the overhead.
The scheuler interrupts and switches tasks, which causes this overhead.

@@

Figure \ref{fig:comparison_msec} shows the execution time of multitasking with the cyclic handler.
A lower limit of the cyclic time is one msec due to the specification of TOPPERS/HRP2, the used RTOS.
More than 10 msec do not be evaluated in this paper because it is thought the larger cyclic time influences applications.
Each execution time of cyclic period is the same as the others.
That is because the overhead of switching tasks (about 3 $\mu$sec) is small in comparison with the execution time.
This results also shows the overhead becomes smaller as the cyclic time is larger, because the overhead depends on the number of switching tasks.
The smaller cyclic time is better in multitasking due to concurrent and/or parallel processing.

@@

\section{Related Work}
\label{sec:Related work}

>>>>

%        File: thesis.tex
%     Created: Tue Feb 16 06:00 PM 2016 J
% Last Change: Tue Feb 16 06:00 PM 2016 J
%
\input{../EMSOFT2016/dummy/dummy.tex}
\documentclass[submit]{ipsj_v2/UTF8/ipsj}
%\documentclass{ipsj}

@@

%\etitle{mruby Bytecode Loader Using Bluetooth in Multi-VM Environment}

\affiliate{OU}{大阪大学基礎工学部\\
School of Engineering Science, Osaka University}

\affiliate{OKUMA}{オークマ株式会社\\
OKUMA Corporation}

\affiliate{GOU}{大阪大学基礎工学研究科\\
Graduate School of Engineering Science, Osaka University}

\author{山本 拓朗}{Takuro Yamamoto}{OU}%[joho.taro@ipsj.or.jp]
\author{大山 博司}{Hiroshi Oyama}{OKUMA}%
\author{安積 卓也}{Takuya Azumi}{GOU}%[gakkai.jiro@ipsj.or.jp]

@@

% \begin{jkeyword}
% 情報処理学会論文誌ジャーナル，\LaTeX，スタイルファイル，べからず集
% \end{jkeyword}

% \begin{eabstract}
% This document is a guide to prepare a draft for submitting to IPSJ
% Journal, and the final camera-ready manuscript of a paper to appear in
% IPSJ Journal, using {\LaTeX} and special style files.  Since this
% document itself is produced with the style files, it will help you to
% refer its source file which is distributed with the style files.
% \end{eabstract}
%
% \begin{ekeyword}
% IPSJ Journal, \LaTeX, style files, ``Dos and Don'ts'' list
% \end{ekeyword}

@@

本論文における貢献は次のとおりである．
\begin{itemize}
        \item　mrubyバイトコードローダの実装により，mruby on TECSを使ったソフトウェア開発における作業効率を向上させる．
        \item　RiteVMスケジューラの実装により，複数のmrubyプログラムを効率的に並行・並列動作させる．
        \item　コンポーネントベース開発の利点を明らかにする．
\end{itemize}
% Contribution: The proposed framework gives the contribution in the following points.
% 1) To improve the software development efficiency:
% Developers do not need to rewrite a storage/ROM device and also to restart an OS. 
% The loader supports the countinuous loading, which saves the Bluetooth setup time. 
% Therefore, the mruby bytecode loader using Bluetooth helps developers develop software.
% 2) To effectively execute multiple mruby programs in concurrent or/and parallel: 
% Developers can execute multiple tasks without the knowledge of an RTOS because the RiteVM sceduler switches tasks cyclically.
% In additon, synchronization of multiple tasks is also implemented.
% 3) To focus on the benefits of component-based developments:
% This paper shows the specific examples for the benefits of component-based development.

本論文の構成を次に述べる．
まず\ref{sec:Background}章で，提案フレームワークのベースであるmruby on TECSについて述べる．
\ref{sec:Design and Implementation}章では，提案フレームワークの設計と実装について述べる．
\ref{sec:Evaluation}章では，ローダを使用した場合の作業効率の比較，RiteVMを使用した場合の実行時間のオーバヘッドを評価する．
\ref{sec:Related Work}章では，関連研究について述べる．
最後に，\ref{sec:Conclusion}章で本論文をまとめる．

@@

\section{既存フレームワーク}
\label{sec:Background}
図\ref{fig:proposed}に提案フレームワークのシステムモデルを示す．
RiteVMとmrubyライブラリを含むプラットフォーム部分は，はじめにターゲットデバイス上で起動されている．
mrubyアプリケーションのバイトコードは，ホストからターゲットデバイスへと転送される．
それぞれバイトコードは，RiteVMに割り当てられ，並行して起動される．

@@

\subsubsection{コンポーネントモデル}
図\ref{fig:component}にコンポーネント図の例を示す．
TECSでは，インスタンス化されたコンポーネントはセル({\it cell})と呼ばれ，受け口({\it entry})，呼び口({\it call})，属性，内部変数を持つ．
受け口は自身の機能を提供するインタフェースで，呼び口は他のセルの機能を利用するためのインタフェースである．
セルは複数の受け口や呼び口を持つことができる．
セルの提供する関数は，C言語で実装される．

受け口と呼び口の型は，セルの機能を使うためのインタフェースであるシグニチャによって定義される．
セルの呼び口は，同じシグニチャを持つ他のセルの受け口と結合できる．

セルの型は，セルタイプと呼ばれ，受け口，呼び口，属性，内部変数の組を定義している．

@@

\subsubsection{コンポーネント記述}
TECSのコンポーネント記述は，シグニチャ記述，セルタイプ記述，組上げ記述の3つに分類され，.cdlファイルに記述する．
図\ref{fig:component}に示されるコンポーネント記述について次に述べる．
\begin{description}
    \item[{\bf シグニチャ記述}]\mbox{}\\
        シグニチャ記述は，セルのインタフェースを定義する．
        {\it signature}キーワードに続けて，シグニチャ名(接頭辞``s")を記述する．
        図\ref{signature}のようにシグニチャsMotorを定義することができる.
        TECSでは，インタフェースの定義を明確にするために，入力と出力にはそれぞれ，[in]と[out]という指定子がつけられる．

\begin{figure}[t]
\centering
\begin{lstlisting}
signature sMotor{
    int32_t getCounts( void );
    ER resetCounts( void );
    ER setPower( [in]int power );
    ER stop( [in] bool_t brake );
    ER rotate( [in] int degrees, [in] uint32_t speed_abs,
              [in] bool_t blocking );
    void initializePort( [in]int32_t type );
};
\end{lstlisting}
\caption{Signature Description}
\label{signature}
\end{figure}

    \item[{\bf セルタイプ記述}]\mbox{}\\
        セルタイプ記述は，受け口，呼び口，属性，内部変数を用いてセルタイプを定義する．
        {\it celltype}キーワードに続けて，セルタイプ名(接頭辞``t")を記述する．
        図\ref{celltype}に示すように，受け口は，{\it entry}キーワードに続けて，シグニチャ名, 受け口名を記述する．
        同様にして，呼び口も定義できる．
        属性と変数は，それぞれ{\it attr}，{\it var}キーワードを用いて列挙する．

\begin{figure}[t]
\centering
\begin{lstlisting}
celltype tCaller{
    call sMotor cMotor;
};
celltype tMotor{
    entry sMotor eMotor;
    attr{
        int32_t attr = 100;
    };
    var{
        int32_t var;
    };
};
\end{lstlisting}
\caption{Celltype Description}
\label{celltype}
\end{figure}

    \item[{\bf 組上げ記述}]\mbox{}\\
        組上げ記述は，セルをインスタンス化し，セル同士を結合する．
        {\it cell}キーワードに続けて，セルタイプ名，セル名を記述する．
        呼び口名，``="，結合先の受け口名の順に記述し，セル同士を結合する．
        図\ref{build}の例では，セルCallerの呼び口cMotorと，セルMotorの受け口eMotorが接続されている．

\begin{figure}[t]
\centering
\begin{lstlisting}
cell tMotor Motor{
};
cell tCaller Caller{
    cMotor = Motor.eMotor;
};
\end{lstlisting}
\caption{Build Description}
\label{build}
\end{figure}

@@

\subsection{mruby on TECS}
\label{sec:mruby on TECS}
mruby on TECSは，mruby(軽量Ruby)を用いた組込みソフトウェアのコンポーネントベース開発が可能なフレームワークである．

\subsubsection{システムモデル}
図\ref{fig:mrubyontecs}にmruby on TECSのシステムモデルを示す．
mrubyバイトコードはそれぞれ，コンポーネント化されたRTOSのタスクとして，RiteVM上で実行される．
TECSのコンポーネントは，モータドライバやセンサドライバのような組込みドライバをサポートしている．

mruby-TECSブリッジによって，mrubyプログラムからC言語の関数を呼び出すことができる．
mruby-TECSブリッジについては，以下で詳しく説明する．

本研究では，RTOSとして，TOPPERS/HRP2 \cite{url:HRP2}, \cite{par:hr-tecs}を使用した．
TOPPERS/HRP2は，μITRON\cite{par:microITRON}をベースにしたメモリ保護機能を持つRTOSである．
しかし，TECSは，TOPPERS/HRP2だけでなく，OSEK \cite{par:OSEK}やTOPPERS/ASP \cite{url:ASP}といったRTOSにも対応しているので，mruby on TECSはRTOSに依存しない．

@@

\subsubsection{mruby-TECSブリッジ}
mrubyとC言語には，実行時間に大きな差がある．
\cite{par:mrubyonTECS}によると, mrubyプログラムは，C言語より実行時間が数百倍遅い．
これによって，mrubyプログラムをすべてアプリケーションに使用することは難しい．
mrubyを使用することにより，生産性やメンテナンス性を向上させることができる一方で，アクチュエータやセンサを制御するようなアプリケーションの重要な部分をC言語で実装することも必要である．

図\ref{fig:mruby_TECS_bridge}に，モータ制御に対するmruby-TECSブリッジの例を示す．
BridgeMotorの左部分は，mrubyプログラムであり，右部分は，TECSのコンポーネントである．
\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/mruby_TECS_bridge.pdf}
    \caption{mruby-TECS Bridge}
    \label{fig:mruby_TECS_bridge}
\end{figure}

mruby-TECSブリッジは，mrubyプログラムからの呼び出しを受け取るためのセルタイプと，C関数を呼び出すためのmrubyクラスを生成する．
生成されたmruby-TECSブリッジのコードがmrubyのクラスとメソッドを登録する．
これによって，mrubyプログラムからあるメソッドが呼ばれると，mruby-TECSブリッジがTECSコンポーネントで定義されたC関数を呼び出す．

\section{設計と実装}
\label{sec:Design and Implementation}
図\ref{fig:system_model}に提案フレームワークの詳細なシステムモデルを示す．
ホストから転送されたmrubyアプリケーションのバイトコードは，それぞれのRiteVMに実装されたローダで受信され，同期処理により同時に実行される．
RiteVMスケジューラがタスクを周期的に切り替えるため，mrubyアプリケーションは並行動作することができる．
　
\begin{figure}[t]
    \centering
    \includegraphics[width=8cm,clip]{../EMSOFT2016/figure/system_model.pdf}
    \caption{Detailed System Model of the proposed framework}
    \label{fig:system_model}
\end{figure}

\subsection{Bluetoothを用いたmrubyバイトコードローダ}
\label{sec:mruby bytecode loader using Bluetooth}
この章では，Bluetoothを用いたmrubyバイトコードローダについて述べる．
現状のmruby on TECSでは，mrubyバイトコードをプラットフォームに組み込んでいるため，mrubyプログラムを修正する度にプラットフォーム部分をもう一度コンパイル・リンクし，SDカードやROMといったデバイスに再度書き込み，OSを再起動する必要がある．
このような作業を繰り返すことで，作業効率は悪くなる．
提案フレームワークでは，プラットフォーム部分をコンパイル・リンクし記憶デバイスに書き込むのは，はじめの一度だけで良いため，作業効率を向上させることができる．

@@

\subsubsection{ローダを実装したRiteVMのコンポーネント}
提案フレームワークでは，mrubyバイトコードローダはTECSのコンポーネントとして提供されている．
このコンポーネントは，RiteVMのコンポーネントの拡張として実装されている．(詳細については\cite{par:mrubyonTECS}を参照)
このコンポーネントでは，転送されたバイトコードの受信に加え，RiteVMのコンフィグレーションが行われる．

@@

\end{lstlisting}
\caption{Main code for RiteVM with mruby bytecode loader using Bluetooth}
\label{maincode_mrubybluetooth}
\end{figure}
\subsection{マルチタスク処理}
\label{sec:Multitask}
この章では，提案フレームワークでのマルチタスク処理の設計にちて述べる．
mruby on TECSは，マルチタスクに対応しているが，開発者がRTOSの知識を熟知している必要がある．

@@

\subsubsection{RiteVMスケジューラ}
RiteVMスケジューラは，周期ハンドラであり，μITRONのサービスコールである{\it rotateReadyQueue}を呼び出す．
{\it rotateReadyQueue}は，同優先度のタスクの実行を切り替える関数である．
RiteVMスケジューラの設計を図\ref{fig:rotateReadyQueue}に示す．

@@

\subsubsection{RiteVMスケジューラのコンポーネント}
図\ref{fig:cyclic_handler}にRiteVMスケジューラのコンポーネント図を示す．
RiteVMスケジューラのコンポーネントは，CyclicHandlerとCyclicMainから構成される．
CyclicHandlerセルは，μITRONの周期ハンドラの設定を行う．
μITRONの周期ハンドラは，\cite{par:microITRON}に詳しく書かれている．

@@

図\ref{build_cyclic_handler}に，図\ref{fig:cyclic_handler}に示されるコンポーネントの組上げ記述を示す．

@@

\subsubsection{mrubyアプリケーションの同期機構}
複数のmrubyアプリケーションの起動を同期させるために，提案フレームワークでは，同期処理の手法としてイベントフラグを適用した．
タスクはそれぞれ0x01(01)，0x02(10)のパターンをセットし，待ちパターン0x3(11)でAND待ちする．
この機構は，2つ以上のタスクの場合でも適用できる．
例えば，タスクが4つの場合，タスクはそれぞれ，0x01(0001), 0x02(0010), 0x04 (0100), 0x08 (1000)のパターンをセットし，待ちパターン0x0f(1111) でAND待ちする．(図\ref{fig:Eventflag} (A))

@@

\subsection{コンポーネントベース開発の利点}
この章では，コンポーネントベース開発の利点を使った設計について述べる．

@@

\section{評価実験}
\label{sec:Evaluation}
この章では，評価実験の結果と考察について述べる．
提案フレームワークの利点を分析するため，以下の評価を行った．
\begin{itemize}
%    \item プラットフォームの起動時間
    \item 転送されるmrubyバイトコードのサイズと処理時間およびコンパイル時間
    \item シングルタスク，コルーチン，マルチタスクの実行時間
    \item 周期時間によるオーバヘッド
%   \item mrubyアプリケーションの同期
    \item コンポーネントベース開発を利用したコードサイズ 
\end{itemize}

% These evaluations are performed in order to indicate that an mruby bytecode loader improves the software development efficiency, and that the proposed multitask processing effectively executes compared with singletasking or co-routine, and also the overhead of the cyclic period.
本評価実験は，ターゲットデバイスとして，LEGO MINDSTORMS EV3 (300MHz ARM9-based Sitara AM1808 system-on-a-chip)を使用し，コンパイラは，gcc 4.9.3 -O2 および mruby version 1.2.0である．

表\ref{tab:size_and_time}に，mrubyアプリケーションのバイトコードと，mrubyライブラリとアプリケーションのバイトコードのサイズ，ロードにかかる処理時間，コンパイル時間の比較を示す．
0バイトのバイトコードを処理した場合にかかるロード処理時間のオーバヘッドは，50.933 msecである．
同様に0バイトのバイトコードをコンパイルした場合にかかるコンパイルのオーバヘッドは，46.9 msecである．
mrubyアプリケーションのみのバイトコードの方が，ライブラリを含めた場合よりも，すべての点で優れている．
これは，RiteVMひとつの評価であるため，RiteVMの数が増えるにつれ，この差は広がっていく．
さらに，この設計では，SDカードやROMを接続したり，OSを再起動する手間も省くことができる．
以上より，提案フレームワークは，既存フレームワークより作業効率を向上させることができる．

% The size, load process time, and compilation time for mruby application and mruby application including library is shown in Table \ref{tab:size_and_time}.
% The overhead of load processing is 50.933 msec, which it takes to load a zero bytes bytecode.
% Similarly, the overhead of compilation is 46.9 msec, which it takes to compile a zero bytes program.
% The mruby application bytecode is smaler and faster than that of including mruby libraries in all terms.
% The difference becomes larger as the number of RiteVMs increases because it takes the 50 msec overhead per one RiteVM. 
% %In the proposed framework, developers send only mruby application and prepare mruby library in advance.
% %Therefore, the design can send the bytecode faster.
% In addition, the design can save the time to connect a storage/ROM device to restart an OS.
% These advantages lead improvement of the software development efficiency.

@@

次に，図\ref{fig:comparison_s_c_m}に，シングルタスク，コルーチン，マルチタスクの実行時間を示す．
評価用のアプリケーションとして，100,000回ループするプログラムを適用した．
マルチタスクおよびコルーチンは，2つのタスクでそれぞれ50,000回ループするプログラムを用いた．
マルチタスクの実行時間は，シングルタスクよりも遅いが，コルーチンより速い．
この結果から，提案フレームワークのマルチタスク処理は，コルーチンより有効である．
RiteVMスケジューラのオーバヘッドは，約5\%である．
タスクの切り替えそのものにかかる時間は，約3μsecである．
RiteVMスケジューラは，周期的に割り込み，タスクの切り替えの処理を行うため，これがオーバヘッドとなっている．

% The comparison of the application execution time with singletasking, co-routine, and multitasking is shown in Figure \ref{fig:comparison_s_c_m}.
% The 100,000 times loop program is used as mruby application for evaluation of execution time.
% In detail, the singletask program loops 100,000 times, and the multitask and co-routine programs loop 50,000 times in each task.
% In Figure \ref{fig:comparison_s_c_m}, the cyclic time of the cyclic handler for multitasking is one msec.
% While multitask's execution time is slower than singletask's, co-routine takes more time than multitask to execute an mruby application.
% This result shows the proposed design is superior to co-routine in terms of execution time.
% The overhead of multitask is about 5 \% ($(multitask-singletask)/multitask$).
% Switching tasks' overhead is also evaluated, that is about three $\mu$sec on average.
% %The number of switching tasks is from 500 to 600 because singletasking process in Figure \ref{fig:comparison_s_c_m} takes about 540 msec.
% %Therefore, the overhead of multitasking is 0.5 \% or less, and multitasking can be used without the overhead.
% The scheuler interrupts and switches tasks, which causes this overhead.

@@

図\ref{fig:comparison_msec}は，RiteVMスケジューラの周期時間ごとの実行時間を示している．
この評価では，TOPPERS/HRP2の仕様により周期時間の下限を1msecとした．
また，周期時間が大きいとアプリケーションに影響を与える可能性があるため，上限を10msecとした．
周期時間が大きいほど，タスクの切り替え回数は減少するため，実行時間は小さくなることが分かる．
1msecと10msecでは，約3\%減少している．
1msecと8msecでは，約1\%減少している．
タスクを並行動作させるには，より小さい周期時間の方が好まれるが，オーバヘッドは大きくなる．
しかし，周期時間によるオーバヘッドは，誤差に埋もれるほど小さい．

% Figure \ref{fig:comparison_msec} shows the execution time of multitasking with the cyclic handler.
% A lower limit of the cyclic time is one msec due to the specification of TOPPERS/HRP2, the used RTOS.
% More than 10 msec do not be evaluated in this paper because it is thought the larger cyclic time influences applications.
% Each execution time of cyclic period is the same as the others.
% That is because the overhead of switching tasks (about 3 $\mu$sec) is small in comparison with the execution time.
% This results also shows the overhead becomes smaller as the cyclic time is larger, because the overhead depends on the number of switching tasks.
% The smaller cyclic time is better in multitasking due to concurrent and/or parallel processing.

@@

\section{関連研究}
\label{sec:Related Work}
